---
title: "Clustering 3K PBMCs with Seurat"
author: "Kevin Rue-Albrecht"
format: html
---

## Libraries

```{r}
#| message: false
#| warning: false
library(cowplot)
library(DropletUtils)
library(scater)
library(scuttle)
```

## Data

### Data download

To create a parallel tutorial to the existing tutorial [Clustering 3K PBMCs with Seurat](https://training.galaxyproject.org/training-material/topics/single-cell/tutorials/scrna-seurat-pbmc3k/tutorial.html), input data files were downloaded from these Zenodo links:

```
https://zenodo.org/record/3581213/files/genes.tsv
https://zenodo.org/record/3581213/files/barcodes.tsv
https://zenodo.org/record/3581213/files/matrix.mtx
```

### SingleCellExperiment object

Import the triplet of files (`matrix.mtx`, `barcodes.tsv`, `genes.tsv`) under `data/`.

```{r}
sce <- DropletUtils::read10xCounts(samples = "data")
```

> A Galaxy tool wrapper exists, [DropletUtils Read10x](https://usegalaxy.eu/?tool_id=toolshed.g2.bx.psu.edu%2Frepos%2Febi-gxa%2Fdropletutils_read_10x%2Fdropletutils_read_10x%2F1.0.4%2Bgalaxy0&version=latest), but produces an `rdata` file which needs to be manually re-typed to `rdata.sce` or `loom` depending on the downstream steps.

Inspect the `SingleCellExperiment` object.

```{r}
print(sce)
```

> New Galaxy tool wrapper needed here, simply to run `print(sce)` into a `txt` file for inspection.

## Preprocessing

### Quality control

Identify mitochondrial genes.

```{r}
mt_gene_ids <- rownames(sce)[grep('^MT-', rowData(sce)[["Symbol"]])]
mt_gene_ids
```

> New Galaxy Tool needed.
> The [Single-cell quality control with scater](https://training.galaxyproject.org/training-material/topics/single-cell/tutorials/scrna-scater-qc/tutorial.html) tutorial provides a precomputed file of mitochondrial genes.

```{r}
gene_subsets <- list()
gene_subsets[["MT"]] <- mt_gene_ids
sce <- scuttle::addPerCellQCMetrics(
	x = sce,
	subsets = gene_subsets
)
```

> A Galaxy Tool wrapper exists, [Scater: calculate QC metrics](https://usegalaxy.eu/?tool_id=toolshed.g2.bx.psu.edu%2Frepos%2Fiuc%2Fscater_create_qcmetric_ready_sce%2Fscater_create_qcmetric_ready_sce%2F1.22.0&version=latest), but requires the expression matrix in tabular format.

### Filtering of low-quality cells

Visualise QC Metrics.

> Note: In contrast to Seurat, scater can only plot one QC metric at a time.

```{r}
#| fig-width: 6
#| fig-height: 3
plot_list <- list()
plot_list[["detected"]] <- scater::plotColData(
	object = sce,
	y = "detected"
)
plot_list[["sum"]] <- scater::plotColData(
	object = sce,
	y = "sum"
)
plot_list[["subsets_MT_percent"]] <- scater::plotColData(
	object = sce,
	y = "subsets_MT_percent"
)
cowplot::plot_grid(plotlist = plot_list, nrow = 1)
```


Filter Out Low Quality Cells

```{r}
keep_cells <- sce$detected >= 200 & sce$detected <= 2500 & sce$subsets_MT_percent <= 5
sce <- sce[keep_cells, ]
```

